---
title: "Fitting state-space spawner-recruitment models to simulated data with Stan"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    df_print: paged
    number_sections: yes
    theme: cerulean
    toc: yes
    toc_float: yes
editor_options:
  chunk_output_type: inline
---

#### Background

This R Markdown document illustrates how to simulate some age-structured spawner recruitment data and then fit a Bayesian state-space spawner recruitment model to it using Stan via [Rstan](https://cran.r-project.org/web/packages/rstan/vignettes/rstan.html). 
The state-space spawner-recruitment model used here is adapted from the model described in [Fleischman et al. 2013](https://cdnsciencepub.com/doi/full/10.1139/cjfas-2012-0112).

First load required libraries and function to simulate spawner-recruitment data.
```{r housekeeping, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rstan)
source("functions.R")
```

#### Simulate some data!

Next, simulate some spawner-recruitment data assuming a Ricker type spawner recruitment relationship, and convert to dataframe.
We need to remove the first `a_max` years of data from this simulation, because recruits do not have observations of spawners from their brood year to be linked to (i.e. these S-R years are not "fully observed").
See `process()` in `functions.R` for details.
Note that if you change `a_min` and `a_max` you may need to update `mat` so its length equals the number of age classes (i.e. `n_a`). 

```{r single simulaiton}
a_min <- 4               #minimum age
a_max <- 7               #maximum age
n_a <- a_max - a_min + 1 #number of age classes  
n_y <- 40                #number of brood years to simulate

sim_sr <- process(n_y = n_y,
                  a_max = a_max,
                  alpha = 4,                   #Ricker alpha (not in log space)
                  beta = 0.0005,               #Ricker beta
                  phi = 0.6,                   #autocorrelation coefficient
                  mat = c(0.2, 0.3, 0.4, 0.1), #maturity schedule, must sum to 1
                  sigma_R = 0.5,               #process variance
                  U_alpha = 7,                 #alpha parameter for beta distribution for annual harvest rate
                  U_beta = 7)                  #beta parameter ' ' 

sim_data <- as.data.frame(sim_sr)[(a_max + 1):n_y, ] #remove years not fully observed
```

The outputs of the function include spawner escapement (`S`), returns (i.e. run) by age (ages 4-7; `N_age.X`), total returns in the year of spawning (`N`, which corresponds to that return year's spawners + catch), and resulting recruitment (`R`) generated by that year's spawners.
This follows the same form as Table 1 in Fleishman et. al. (2013); note that table uses *Esc* and *Return*, where we use *Spawners* (`S`) and *Recruitment* (`R`).

```{r}
head(sim_data)
```

Before we try and fit a state-space model to the data let's just fit a simple linear Ricker model to the data and see what the relationship looks like.

```{r}
lm_Ricker <- lm(log(sim_data$R/sim_data$S)~sim_data$S)
lm_Ricker_a <- lm_Ricker$coefficients[1]
lm_Ricker_b <- lm_Ricker$coefficients[2]

S <- seq(0, 4500, length.out = 100) #vector of spawner abundances to predict relationship

pred_R <- exp(lm_Ricker_a) * S * exp(lm_Ricker_b * S)
sr_pred <- as.data.frame(cbind(S, pred_R))

ggplot(data = sim_data, aes(x = S, y = R)) +
  geom_point() +
  xlim(0, NA) +
  ylim(0, NA) +
  geom_line(data = sr_pred, aes(x = S, y = pred_R), color="black", lwd = 1) +
  theme_bw()

```

The estimates of alpha (`r round(exp(lm_Ricker$coefficients[1]),2)`) and beta (`r round(lm_Ricker$coefficients[2],4)`) are similar to the true values (alpha = 4, beta = 0.0005) but not identical. 


#### Fit a state-space model

Before fitting the model we need to wrangle the simulated data into an appropriate format. First we need to generate a data frame of observed age composition over time, and specify it as integers with the sum of the rows equal to the effective sample size for each year (for simplicity assumed to equal 100 *each year*, which wouldn't be the case if a field program collected different amounts of age data each year):

```{r}
A_obs <- sim_data |>
  mutate(age.4 = N_age.1/N,
         age.5 = N_age.2/N,
         age.6 = N_age.3/N,
         age.7 = N_age.4/N) |>
  select(age.4:age.7)

A_obs <- round(as.matrix(A_obs)*100) #then mult by 100 to scale for sample size 
```

Next we need to generate data frames of spawner and harvest observations; we treat the simulated data as truth then add normally distributed error of 20% of the mean for spawners and 5% for harvest. 
This extra noise arises from not being able to perfectly assess the number of spawners or count up all the catch.  

```{r}
S_obs <- rnorm(length(sim_data$S), 
               sim_data$S,
               0.2*sim_data$S)

H_obs <- rnorm(length(sim_data$S),
               (sim_data$N-sim_data$S),
               0.05*(sim_data$N-sim_data$S))
```

And then specify a list of the data to be read when fitting the model: 

***need to describe why n_R_y is longer than n_y really well here***  
Most assessments start with "fully observed" return years of S, R, and N (then maybe assume some A_obs)? 
We don't have to trim off the first `a_max` years of data like we did when we simulated because the population dynamics are already there in nature. 
Here, we have to add `a_max` unobserved recruitments to the end of our timeseries so we have something to link to the observed spawners.



```{r}
n_y <- length(S_obs) #number of fully observed spawning (i.e. calendar/return) years
n_R_y <- n_y + a_max + 1 #total number of not-fully-observed recruitment (i.e. brood) years needed to link S-R in the model
#n_R_y <- nyrs + A - 1 # number of recruitment years ## should be nyrs + A_max + 1 to get all unobserved states?
#nRyrs <- nyrs + a_max + 1 # other option


stan.data <- list("n_y" = n_y, #number of return years
                  "a_min" = a_min,
                  "a_max" = a_max,
                  "n_a" = n_a,
                  "n_R_y" = n_R_y,
                  "A_obs" = A_obs,
                  "S_obs" = S_obs,
                  "H_obs" = H_obs,
                  "S_CV" = rep(0.2,length(S_obs)),
                  "H_CV" = rep(0.05,length(H_obs)))
```

The actual model is written out as a Stan program (see `SS-SR_AR1.stan`) which is then called when fitting the model using the `stan()` function like so:

```{r}
stan.fit <- stan(file = "SS-SR_AR1.stan",
                 model_name = "SS-SR_AR1",
                 data = stan.data,
                 chains = 2,
                 iter = 500,
                 seed = 42,
                 thin = 1,
                 control = list(adapt_delta = 0.99, max_treedepth = 20))
```

For brevity I have only run two chains for 500 iterations. You will get some warnings that are all worth looking into, but we will ignore for now.

Shinystan is a great tool for interactive model examination and can be called by commenting this code snipit:

```{r}
#shinystan::launch_shinystan(stan.fit) 
```

Save the fitted model and data objects:

```{r}
saveRDS(stan.fit, file="./output/SS-SR_AR1.stan.fit.rds")
saveRDS(stan.data, file="./output/SS-SR_AR1.stan.data.rds")
```

Or read in the objects if working from a previously save set, and extract posteriors:

```{r}
stan.fit <- readRDS("output/SS-SR_AR1.stan.fit.rds")
stan.data <- readRDS("output/SS-SR_AR1.stan.data.rds")
model_pars <- rstan::extract(stan.fit)
```

#### Inference! 

So we have now fitted our state-space model. Lets see how well the posterior samples recover the true underlying parameters we used to simulate the data (red vertical lines):

```{r, message = FALSE}
post_df <- data.frame(
  parameter = factor(rep(c("lnalpha", "beta", "phi"), each=dim(model_pars$lnalpha))),
  posterior = c(model_pars$lnalpha, model_pars$beta, model_pars$phi)
  )

true_para_df <- data.frame(
  parameter = factor(c( "beta", "lnalpha", "phi")),
  value = c(0.0005, log(4), 0.6)
  )

ggplot(post_df, aes(x = posterior)) + 
  geom_histogram() +
  facet_wrap(~ parameter, scales="free") +
  geom_vline(data = true_para_df, aes(xintercept=value), color = "red") +
  theme_bw()
```

And what does the inferred spawner-recruitment relationship look like?. First we need to generate a data frame of spawner abundance and predicted recruitment (with uncertainty) based on the posterior samples:

```{r}
max_samples <- dim(model_pars$lnalpha)

spwn <- exp(model_pars$lnS)
spwn.quant <- apply(spwn, 2, quantile, probs=c(0.05,0.5,0.95))[,1:(nyrs-a_min)]

rec <-exp(model_pars$lnR)
rec.quant <- apply(rec, 2, quantile, probs=c(0.05,0.5,0.95))[,(a_max+1):dim(model_pars$R)[2]] #fully observed R

brood_t <- as.data.frame(cbind(1:(nyrs-a_min),t(spwn.quant), t(rec.quant)))
colnames(brood_t) <- c("BroodYear","S_lwr","S_med","S_upr","R_lwr","R_med","R_upr")

brood_t <- as.data.frame(brood_t)

# SR relationship
spw <- seq(0,max(brood_t[,a_min]),length.out=100)
SR_pred <- matrix(NA,100,max_samples)

for(i in 1:max_samples){
  r <- sample(seq(1,max_samples),1,replace=T)
  a <- model_pars$lnalpha[r]
  b <- model_pars$beta[r]
  SR_pred[,i] <- (exp(a)*spw*exp(-b*spw))
}

SR_pred <- cbind(spw,t(apply(SR_pred,c(1),quantile,probs=c(0.05,0.5,0.95),na.rm=T)))
colnames(SR_pred) <- c("Spawn", "Rec_lwr","Rec_med","Rec_upr")
SR_pred <- as.data.frame(SR_pred)
```

And then plot out relationship:

```{r, warning = FALSE}
ggplot() +
  geom_ribbon(data = SR_pred, aes(x = Spawn, ymin = Rec_lwr, ymax = Rec_upr),
              fill = "grey80", alpha=0.5, linetype=2, colour="gray46") +
  geom_line(data = SR_pred, aes(x = Spawn, y = Rec_med), color="black", size = 1) +
  geom_errorbar(data = brood_t, aes(x= S_med, y = R_med, ymin = R_lwr, ymax = R_upr),
                colour="grey", width=0, size=0.3) +
  geom_errorbarh(data = brood_t, aes(x= S_med, y = R_med, xmin = S_lwr, xmax = S_upr),
                 height=0, colour = "grey", height = 0, size = 0.3) +
  geom_point(data = brood_t, aes(x = S_med, y = R_med, color=BroodYear, width = 0.9), size = 3)+
  coord_cartesian(xlim=c(0, max(brood_t[,4])), ylim=c(0,max(brood_t[,7])), expand = FALSE) +
  scale_colour_viridis_c()+
  xlab("Spawners") +
  ylab("Recruits") +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.key.size = unit(0.4, "cm"),
        legend.title = element_text(size=9),
        legend.text = element_text(size=8))+
  geom_abline(intercept = 0, slope = 1,col="dark grey")
```